---
title: Assess Poisson NMF fits in droplet data
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we compare the quality of the fits obtained from the different
updates (EM, CCD, SCD, with and without extrapolation), and with
different $k$.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center")
```

These are the packages used in the analysis below.

```{r load-pkgs, message=FALSE}
library(fastTopics)
library(ggplot2)
library(cowplot)
source("../code/functions_for_assessing_fits.R")
```

Load the results of running `fit_poisson_nmf` on the droplet data,
with different algorithms, and for various choices of $k$ (the number
of "topics").

```{r load-results}
load("../output/droplet/fits-droplet.RData")
```

This plot shows the improvement in the log-likelihood as the rank,
$k$, is increased. The log-likelihoods are shown relative to the
log-likelihood at $k = 2$. 

```{r plot-loglik-vs-k, fig.height=3, fig.width=3.5}
plot_loglik_vs_rank(fits) +
  theme_cowplot(font_size = 12)
```

Oddly, the likelihood did not improve from $k=4$ to $k=5$.
Otherwise, the likelihood improved with larger rank, $k$.

The next set of plots shows the improvement in the fit over time,
for all choices of $k$, and for each of the three optimization methods
(EM, CCD, SCD), with and without the extrapolation scheme. The quality
of the fit is measured by the log-likelihood relative to the best
log-likelihood that was identified among all methods compared.

```{r plot-loglik, fig.height=6, fig.width=9}
create_progress_plots(dat,fits,"loglik")
```

The second set of plots shows the evolution of Karush-Kuhn-Tucker
(KKT) residuals over time; the KKT residuals should vanish near a
stationary point of the log-likelihood, so looking at the largest KKT
residual can be used to assess how close we are to a solution. Note
that, unlike the likelihood, the residuals of the KKT conditions are
not expected to decrease monotonically over time.

```{r plot-res, fig.height=6, fig.width=9}
create_progress_plots(dat,fits,"res")
```

