---
title: "Identify clusters in droplet data using topic model"
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we perform PCA on the topic proportions to identify clusters in
the droplet data. 

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the packages used in the analysis below, as well as additional
functions that we will use to generate some of the plots.

```{r load-pkgs, message=FALSE}
library(dplyr)
library(fastTopics)
library(ggplot2)
library(cowplot)
source("../code/plots.R")
```

Load data and results
---------------------

Load the droplet data. The UMI counts are not needed for this analysis.

```{r load-data}
load("../data/droplet.RData")
rm(counts)
```

Load the $k = 7$ Poisson NMF model fit.

```{r load-fit}
fit <- readRDS("../output/droplet/rds/fit-droplet-scd-ex-k=7.rds")$fit
```

Identify clusters from principal components
-------------------------------------------

To identify clusters, we begin by plotting PCs computed from the topic
proportions. (Note that only 6 PCs are needed for 7 topics.)

```{r clustering-1, fig.width=8, fig.height=2.5}
p1 <- basic_pca_plot(fit,1:2)
p2 <- basic_pca_plot(fit,3:4)
p3 <- basic_pca_plot(fit,5:6)
plot_grid(p1,p2,p3,nrow = 1,ncol = 3)
```

Some of the structure is more evident from "hexbin" plots showing the
density of the points.

```{r clustering-2, fig.width=8, fig.height=2.5}
bins <- c(0,1,5,10,100,Inf)
p4 <- pca_hexbin_plot(fit,1:2,bins = bins) + guides(fill = "none")
p5 <- pca_hexbin_plot(fit,3:4,bins = bins) + guides(fill = "none")
p6 <- pca_hexbin_plot(fit,5:6,bins = bins) + guides(fill = "none")
plot_grid(p4,p5,p6,nrow = 1,ncol = 3)
```

From these PCA plots, we define 4 clusters, labeled A, Cil, G and
T+N. (The reasoning behind these labels will become clear later.)
Points that do not fit in any of these clusters are assigned to a
"background cluster", labeled U.

```{r clustering-3}
pca <- prcomp(poisson2multinom(fit)$L)$x
x   <- rep("U",nrow(pca))
pc1 <- pca[,1]
pc2 <- pca[,2]
pc6 <- pca[,6]
x[pc2 > -0.15] <- "A"
x[pc1 > 0.3 & pc2 < -0.75] <- "Cil"
x[pc1 <= 0.3 & pc2 >= -0.75 & pc2 < -0.4] <- "T+N"
x[pc6 < -0.05] <- "G"
```

There is additional substructure in cluster A, which is more apparent
in the projection onto the top 2 PCs computed from cluster A only.

```{r clustering-4, fig.width=6, fig.height=3}
rows <- which(x == "A")
fit2 <- select(poisson2multinom(fit),loadings = rows)
p7   <- basic_pca_plot(fit2,1:2)
p8   <- pca_hexbin_plot(fit2,1:2,bins = bins) + guides(fill = "none")
plot_grid(p7,p8)
```

*Add text here.*

```{r clustering-5, fig.width=5.5, fig.height=4}
p9 <- pca_plot(fit2,pcs = 1:2,k = c(2,4,5))
print(p9)
```

We label the three more-or-less distinct subclusters as B, C and H,
and assign the remaining samples to the background cluster.

```{r clustering-6}
pca <- prcomp(fit2$L)$x
y   <- rep("U",nrow(pca))
pc1 <- pca[,1]
pc2 <- pca[,2]
y[pc1 < 0.1] <- "B"
y[pc1 > 0.4 & pc2 < 0.45] <- "C"
y[pc2 > 0.55] <- "H"
x[rows] <- y
```

In summary, we have subdivided the droplet data into 7 subsets,
including a background cluster (U).

```{r clustering-7, fig.width=8, fig.height=2}
colors <- c("royalblue","forestgreen","firebrick","gold","turquoise",
            "darkorange","gainsboro")
p10 <- labeled_pca_plot(fit,1:2,x,colors,font_size = 8)
p11 <- labeled_pca_plot(fit,3:4,x,colors,font_size = 8)
p12 <- labeled_pca_plot(fit,5:6,x,colors,font_size = 8)
plot_grid(p10,p11,p12,nrow = 1,ncol = 3)
```

The clusters identified here correspond well to the Montoro *et al*
(2018) clustering, with some exceptions (e.g., we do not identify an
ionocytes cluster, and the neuroendocrine and tuft cells are included
in the same cluster).

```{r clustering-8}
samples$cluster <- factor(x,c("B","C","H","Cil","T+N","G","U"))
with(samples,table(tissue,cluster))
```

Structure plot
--------------

*Add text here.*

```{r structure-plot, fig.width=8, fig.height=1.5}
set.seed(1)
topic_colors <- c("gold","royalblue","salmon","turquoise","olivedrab",
                  "firebrick","forestgreen")
topics <- c(3,4,5,1,7,2,6)
rows <- sort(c(sample(which(samples$cluster == "B"),800),
               sample(which(samples$cluster == "C"),800),
               which(samples$cluster == "H"),
               which(samples$cluster == "Cil"),
               which(samples$cluster == "T+N"),
               which(samples$cluster == "G"),
			   which(samples$cluster == "U")))
p13 <- structure_plot(select(poisson2multinom(fit),loadings = rows),
                      grouping = samples[rows,"cluster"],
                      topics = topics,colors = topic_colors[topics],
		  			  perplexity = c(70,70,30,50,30,12,50),
                      n = Inf,gap = 30,num_threads = 4,verbose = FALSE)
print(p13)
```

TO DO: Show PCA plots suggesting additional hetergeneity in C cluster.

Save results
------------

Save the clustering of the droplet data to an RDS file.

```{r save-results}
saveRDS(samples,"clustering-droplet.rds")
```
