---
title: "Identify clusters in mixture of FACS-purified PBMC data using topic model"
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we perform PCA on the topic proportions to identify clusters in
the mixture of FACS-purified PBMC data sets.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the packages used in the analysis below, as well as additional
functions that we will use to generate some of the plots.

```{r load-pkgs, message=FALSE}
library(Matrix)
library(dplyr)
library(fastTopics)
library(Rtsne)
library(uwot)
library(ggplot2)
library(cowplot)
source("../code/plots.R")
```

Load the count data.

```{r load-data}
load("../data/pbmc_purified.RData")
```

Load the $K = 6$ Poisson NMF model fit.

```{r load-fit}
fit <- readRDS(file.path("../output/pbmc-purified/rds",
                         "fit-pbmc-purified-scd-ex-k=6.rds"))$fit
```

Identify clusters from principal components
-------------------------------------------

From the PCs, we define clusters for B-cells, CD14+ cells and CD34+
cells. The remaining cells are assigned to the U cluster ("U" for
"unknown").

```{r clustering-1}
pca <- prcomp(poisson2multinom(fit)$L)$x
n   <- nrow(pca)
x   <- rep("U",n)
pc1 <- pca[,1]
pc2 <- pca[,2]
pc3 <- pca[,3]
pc4 <- pca[,4]
pc5 <- pca[,5]
x[pc2 > 0.25] <- "B"
x[pc3 < -0.2 & pc4 < 0.2] <- "CD34+"
x[(pc4 + 0.1)^2 + (pc5 - 0.8)^2 < 0.07] <- "CD14+"
```

Next, we define an NK cells from the top 2 PCs of the unknown cells.

```{r clustering-2}
rows <- which(x == "U")
n    <- length(rows)
fit2 <- select(poisson2multinom(fit),loadings = rows)
pca  <- prcomp(fit2$L)$x
y    <- rep("U",n)
pc1  <- pca[,1]
pc2  <- pca[,2]
y[pc1 < -0.3 & 1.1*pc1 < -pc2 - 0.57] <- "NK"
x[rows] <- y
```

Among the remaining cells, there is a much less distinct cluster of
CD8+ cells, and the rest are labeled as T-cells.

```{r clustering-3}
rows <- which(x == "U")
n    <- length(rows)
fit2 <- select(poisson2multinom(fit),loadings = rows)
pca  <- prcomp(fit2$L)$x
y    <- rep("T",n)
pc1  <- pca[,1]
pc2  <- pca[,2]
y[pc1 < 0.25 & pc2 < -0.15] <- "CD8+"
x[rows] <- y
```

In summary, we have subdivided the cells into 6 subsets. The
substructure is clear when we plot T cells separately from the other
cells.

```{r clustering-5, fig.width=7.5, fig.height=4, message=FALSE}
cluster_colors <- c("dodgerblue","forestgreen","darkmagenta",
                    "red","gray","darkorange")
samples$cluster <- factor(x)
x     <- with(samples,cluster == "T" | cluster == "CD8+")
rows1 <- sample(which(!x))
rows2 <- sample(which(x))
p1 <- pca_plot(select(poisson2multinom(fit),loadings = rows1),
               fill = samples$cluster[rows1]) +
  scale_fill_manual(values = cluster_colors,drop = FALSE) +
  labs(fill = "cluster")
p2 <- pca_plot(select(poisson2multinom(fit),loadings = rows2),
               fill = samples$cluster[rows2,drop = FALSE]) +
  scale_fill_manual(values = cluster_colors,drop = FALSE) +
  labs(fill = "cluster")
p3 <- pca_hexbin_plot(select(poisson2multinom(fit),loadings = rows1),bins = 28)
p4 <- pca_hexbin_plot(select(poisson2multinom(fit),loadings = rows2),bins = 28)
plot_grid(p1,p3,p2,p4)
```

This clustering corresponds closely to the Zheng *et al* (2017) FACS
cell populations:

```{r clustering-6}
with(samples,table(celltype,cluster))
```

This close correspondence is also apparent from PCA plots:

```{r clustering-7, fig.width=8, fig.height=1.75, message=FALSE}
set.seed(1)
facs_colors <- c("dodgerblue",  # B-cells
                 "forestgreen", # CD14+
                 "darkmagenta", # CD34+
                 "firebrick",   # T helper cells
                 "gray",        # NK cells
                 "tomato",      # cytotoxic T-cells
                 "yellow",      # memory T-cells
                 "magenta",     # naive cytotoxic 
                 "darkorange",  # naive T-cells
                 "gold")        # regulatory T-cells
p1 <- pca_plot(select(poisson2multinom(fit),loadings = rows1),
               fill = samples[rows1,"celltype"]) +
  scale_fill_manual(values = facs_colors) +
  labs(fill = "FACS subtype")
p2 <- pca_plot(select(poisson2multinom(fit),loadings = rows2),
               fill = samples[rows2,"celltype"]) +
  scale_fill_manual(values = facs_colors) +
  labs(fill = "FACS subtype")
plot_grid(p1,p2)
```

Structure plot
--------------

The structure plot summarizes the topic proportions in each of these 6
subsets:

```{r structure-plot, fig.width=8, fig.height=1.5}
set.seed(1)
topic_colors <- c("gold","forestgreen","dodgerblue","gray","darkmagenta",
                  "violet")
topics <- c(5,3,2,4,1,6)
rows <- sort(c(sample(which(samples$cluster == "B"),1000),
               sample(which(samples$cluster == "CD14+"),300),
               sample(which(samples$cluster == "CD34+"),500),
               sample(which(samples$cluster == "CD8+"),400),
               sample(which(samples$cluster == "NK"),500),
               sample(which(samples$cluster == "T"),1000)))
p3 <- structure_plot(select(poisson2multinom(fit),loadings = rows),
                     grouping = samples[rows,"cluster"],
                     topics = topics,colors = topic_colors[topics],
   	  			     perplexity = c(70,30,30,30,30,70),
                     n = Inf,gap = 50,num_threads = 4,verbose = FALSE)
print(p3)
```

Analysis of single-cell likelihoods
-----------------------------------

*Add text and code here.*

```{r}
set.seed(5)
rows <- which(with(samples,
                   cluster == "B" |
                   cluster == "CD14+" |
                   cluster == "CD34+"))
rows <- sample(rows,2000)
fit2 <- select(poisson2multinom(fit),loadings = rows)
p3   <- pca_plot(fit2,fill = samples[rows,"cluster"])
```

Run t-SNE, then plot the 2-d embedding.

```{r}
tsne <- Rtsne(fit2$L,dims = 2,pca = FALSE,normalize = FALSE,perplexity = 100,
              theta = 0.1,max_iter = 1000,eta = 200,verbose = TRUE)
tsne$x <- tsne$Y
colnames(tsne$x) <- c("tsne1","tsne2")
p4 <- pca_plot(fit2,out.pca = tsne,fill = samples[rows,"cluster"])
```

Run UMAP, then plot the 2-d embedding.

```{r}
out.umap <- umap(fit2$L,n_neighbors = 30,metric = "euclidean",n_epochs = 1000,
                 min_dist = 0.1,scale = "none",learning_rate = 1,
                 verbose = TRUE)
out.umap <- list(x = out.umap)
colnames(out.umap$x) <- c("umap1","umap2")
p5 <- pca_plot(fit2,out.pca = out.umap,fill = samples[rows,"cluster"])
plot_grid(p3,p4,p5,nrow = 1)
```

TO DO:

+ Compare inter-cluster and inter-topic total variation distances, as
  well as total variation distances between FACS cell populations.
 
+ Create scatterplots comparing single-cell likelihoods using a "hard"
  clustering (either my own clustering, or the FACS-sorted cell
  populations) and using topic model.

+ Add example illustrating how t-SNE and UMAP accentuate clustering
  (see code above).
