---
title: "Evaluation of DE analysis with more than two topics"
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we expand the evaluation of the new topic-model-based DE analysis
methods to consider the case of more than two topics. We can no longer
compare to DESeq2 because DESeq2 is designed to compare gene
expression across two groups. Still, we can closely examine the
benefit of accounting for uncertainty, and in so doing we will see
that the benefits become more pronounced because our analysis factors
in correlations in the estimates across the different topics.

When comparing multiple groups, clusters or topics it is lessclear
what it means for a gene to be "differentially expressed". So to
simplify the evaluation we simulate the data so that each gene is
differentially expressed in at most one topic.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.show = "hold",fig.align = "center",dpi = 120)
```

Load the packages used in the analysis below, and some additional
functions for simulating the data.

```{r load-pkgs, message=FALSE, warning=FALSE}
library(Matrix)
library(fastTopics)
library(ggplot2)
library(cowplot)
source("../code/de_analysis_functions.R")
```

Simulate UMI counts
-------------------

We begin by simulating UMI count data from a rank-6 Poisson NMF model
such that roughly half of the genes have the same expression rate
across all topics, and the other half have a different rate of
expression in exactly one out of the six topics. The details of the
simulation are similar to the two-topic case.

```{r simulate-counts}
set.seed(1)
m <- 10000
k <- 6
dat <- simulate_manytopic_umi_data(m = m,k = k)
X <- dat$X
F <- dat$F
L <- dat$L
```

Fit multinomial topic model to UMI counts
-----------------------------------------

Now we fit a multinomial topic model to the simulated UMI count data.
To simplify evaluation, we assume that the topic proportions are
known, and fix them to their ground-truth values. In this way, the
only error that can arise is in the estimates of the expression rates
$f_{ij}$.

```{r fit-poisson-nmf, warning=FALSE, message=FALSE, cache=TRUE}
fit0 <- init_poisson_nmf(X,F = dat$F,L = with(dat,s*L))
fit <- fit_poisson_nmf(X,fit0 = fit0,numiter = 40,method = "scd",
                       update.loadings = NULL,verbose = "none")
fit <- poisson2multinom(fit)
summary(fit)
```

DE analysis with and without shrinkage
--------------------------------------

First we perform a DE analysis *without* shrinking the LFC estimates:

```{r de-noshrink, message=FALSE, cache=TRUE}
set.seed(1)
de.noshrink <- de_analysis(fit,X,shrink.method = "none",
                           control = list(ns = 10000,nc = 2))
```

Next we perform a second DE analysis using adaptive shrinkage to
shrink the LFC estimates:

```{r de-with-ash, message=FALSE, cache=TRUE}
set.seed(1)
de <- de_analysis(fit,X,shrink.method = "ash",control = list(ns = 1e4,nc = 2))
```

*Add text here.*

```{r shrink-vs-noshrink-scatterplot, fig.height=2.75, fig.width=5}
pdat <- data.frame(noshrink  = c(de.noshrink$postmean),
                   shrink    = c(de$postmean),
                   log10mean = rep(log10(de$f0),k))
p1 <- ggplot(pdat,aes(x = noshrink,y = shrink,fill = log10mean)) +
  geom_point(shape = 21,color = "white",size = 2) +
  geom_abline(intercept = 0,slope = 1,color = "black",linetype = "dotted") +
  scale_fill_gradient2(low = "deepskyblue",mid = "gold",high = "orangered",
                       midpoint = -4) +
  labs(x = "original LFC estimates",
       y = "shrunken LFC estimates") +
  theme_cowplot()
print(p1)
```

*Add text here.*

```{r shrink-vs-noshrink-zscores, fig.height=2.75, fig.width=8, warning=FALSE}
nonzero_lfc <- matrix(FALSE,m,k)
for (i in 1:m) {
  y <- dat$F[i,]
  if (max(y) - min(y) > 1e-8) {
    j <- which.max((y - mean(y))^2)
    nonzero_lfc[i,j] <- TRUE
  }
}
pdat <- data.frame(noshrink = clamp(c(de.noshrink$z),-4,+4),
                   shrink   = clamp(c(de$z),-4,+4),
                   de       = factor(c(nonzero_lfc)))
p2 <- ggplot(pdat,aes(x = noshrink,color = de,fill = de)) +
  geom_histogram(bins = 64) +
  scale_color_manual(values = c("darkorange","darkblue")) +
  scale_fill_manual(values = c("darkorange","darkblue")) +
  labs(x = "z-score",y = "genes",title = "without shrinkage") +
  theme_cowplot()
p3 <- ggplot(pdat,aes(x = shrink,color = de,fill = de)) +
  geom_histogram(bins = 64) +
  scale_color_manual(values = c("darkorange","darkblue")) +
  scale_fill_manual(values = c("darkorange","darkblue")) +
  labs(x = "z-score",y = "genes",title = "with shrinkage") +
  theme_cowplot()
print(plot_grid(p2,p3))
```

*Add text here.*

```{r shrink-vs-noshrink-pvalues, fig.height=2.75, fig.width=8, warning=FALSE}
pdat <- data.frame(noshrink = 10^(-c(de.noshrink$lpval)),
                   shrink   = 10^(-c(de$lpval)),
                   de       = factor(c(nonzero_lfc)))
p4 <- ggplot(pdat,aes(x = noshrink,color = de,fill = de)) +
  geom_histogram(bins = 64) +
  scale_color_manual(values = c("darkorange","darkblue")) +
  scale_fill_manual(values = c("darkorange","darkblue")) +
  labs(x = "p-value",y = "genes",title = "without shrinkage") +
  theme_cowplot()
p5 <- ggplot(pdat,aes(x = shrink,color = de,fill = de)) +
  geom_histogram(bins = 64) +
  scale_color_manual(values = c("darkorange","darkblue")) +
  scale_fill_manual(values = c("darkorange","darkblue")) +
  labs(x = "p-value",y = "genes",title = "with shrinkage") +
  theme_cowplot()
print(plot_grid(p4,p5))
```

*Add text here.*

Quantifying uncertainty in the LFC estimates
--------------------------------------------

*Add text here.*

```{r point-vs-shrunken-scatterplot, fig.width=8, fig.height=2.5}
pdat <- data.frame(est       = c(de$est),
                   shrink    = c(de$postmean),
				   log10mean = rep(log10(de$f0),k))
p6 <- ggplot(pdat,aes(x = est,y = shrink,fill = log10mean)) +
  geom_point(shape = 21,color = "white",size = 2) +
  geom_abline(intercept = 0,slope = 1,color = "black",linetype = "dotted") +
  scale_fill_gradient2(low = "deepskyblue",mid = "gold",high = "orangered",
                       midpoint = -4) +
  labs(x = "point estimates",
       y = "shrunken estimates") +
  theme_cowplot()
print(p6)
```

*Add text here.*

```{r kl-vs-fasttopics-1, fig.width=8, fig.height=2.5}
D     <- fastTopics:::min_kl_poisson(fit$F)
i     <- which(de$z > 0)
pdat1 <- data.frame(kl = pmin(D[i],5e-6),de = factor(nonzero_lfc[i]))
pdat2 <- data.frame(z = clamp(de$z[i],-2,+2),de = factor(nonzero_lfc[i]))
p7 <- ggplot(pdat1,aes(x = kl,color = de,fill = de)) +
  geom_histogram(bins = 64) +
  scale_color_manual(values = c("darkorange","darkblue")) +
  scale_fill_manual(values = c("darkorange","darkblue")) +
  labs(x = "K-L divergence",y = "genes") +
  theme_cowplot()
p8 <- ggplot(pdat2,aes(x = z,color = de,fill = de)) +
  geom_histogram(bins = 64) +
  scale_color_manual(values = c("darkorange","darkblue")) +
  scale_fill_manual(values = c("darkorange","darkblue")) +
  labs(x = "z-score",y = "genes") +
  theme_cowplot()
print(plot_grid(p7,p8))
```

*Add text here.*

```{r kl-vs-fasttopics-2, fig.width=8, fig.height=2.5}
p9  <- p7 + scale_x_continuous(limits = 1e-6 * c(0.5,5.1))
p10 <- p8 + scale_x_continuous(limits = c(0.05,2.1))
print(plot_grid(p9,p10))
```
