---
title: "Comparison of fastTopics DE analysis with MAST and DESeq2"
author: Peter Carbonetto
output: workflowr::wflow_html
---

This is a more systematic comparison of the new fastTopics DE analysis
with MAST and DESeq2 based on an
[initial smaller simulation](de_analysis_detailed_look.html). The
results were generated using the `run_sims.R` script, and here we
summarize the results of these simulations.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.show = "hold",fig.align = "center",dpi = 120)
```

Load the packages needed for this analysis, and some additional
functions used to compile the results and generate the plots.

```{r load-pkgs, message=FALSE}
library(Matrix)
library(Seurat)
library(DESeq2)
library(MAST)
library(fastTopics)
library(ggplot2)
library(cowplot)
source("../code/de_analysis_functions.R")
```

Load the results of the simulations.

```{r load-sims}
load("../output/sims/sims-k=2-alpha=0.01.RData")
```

These were the R packages used to run the simulations:

```{r session-info}
res$session.info
```

Since we are comparing with MAST and DESeq2, two methods that do not
allow for partial membership to groups, we have simulated count data
from a topic model in which the true topic proportions are all 0 or
1, or mostly very close to 0 or 1.

```{r check-topic-props}
res["session.info"] <- NULL
x <- combine_sim_res(res,function (x) apply(x$dat$L,1,max))
mean(x > 0.99)
```

Before comparing the methods, first we assess accuracy of the Monte
Carlo computations by comparing the estimates from two independent
MCMC simulations. The estimates from the two independent simulations
are largely consistent.

```{r mcmc-scatterplots, fig.height=3, fig.width=6}
pdat <- data.frame(lfc1 = combine_sim_res(res,function (x) x$de1$postmean[,2]),
                   lfc2 = combine_sim_res(res,function (x) x$de2$postmean[,2]),
                   z1   = combine_sim_res(res,function (x) x$de1$z[,2]),
		   z2   = combine_sim_res(res,function (x) x$de2$z[,2]))
p1 <- ggplot(pdat,aes(x = lfc1,y = lfc2)) +
  geom_point(color = "mediumblue",shape = 4,size = 0.75) +
  geom_abline(intercept = 0,slope = 1,linetype = "dotted") +
  labs(x = "first posterior mean",y = "second posterior mean") +
  theme_cowplot(font_size = 12)
p2 <- ggplot(pdat,aes(x = z1,y = z2)) +
  geom_point(color = "mediumblue",shape = 4,size = 0.75) +
  geom_abline(intercept = 0,slope = 1,linetype = "dotted") +
  labs(x = "first z-score estimate",y = "second z-score estimate") +
  xlim(-45,42) + 
  ylim(-45,42) + 
  theme_cowplot(font_size = 12)
plot_grid(p1,p2)
```

```{r mcmc-scatterplots-for-paper, echo=FALSE}
ggsave("../plots/mcmc_scatterplots_sims_k=2_alpha=0.01.png",
       plot_grid(p1,p2),height = 3,width = 6,dpi = 600)
```

Now we compare LFC estimates for the second topic:

```{r lfc-scatterplots, fig.height=2.5, fig.width=7}
j  <- paste0("g",1:10000)
lfc.deseq <- combine_sim_res(res,function (x) x$deseq$log2FoldChange)
lfc.fasttopics <- combine_sim_res(res,function (x) x$de1$postmean[,2])
lfc.mast <- combine_sim_res(res,function(x)x$mast[j,"avg_log2FC"])
z.deseq <- combine_sim_res(res,function (x) with(x$deseq,log2FoldChange/lfcSE))
z.fasttopics <- combine_sim_res(res,function (x) x$de1$z[,2])
pdat <- data.frame(lfc.deseq      = lfc.deseq,
                   lfc.fasttopics = lfc.fasttopics,
                   lfc.mast       = clamp(lfc.mast,-6,+6),
                   z.deseq        = z.deseq,
                   z.fasttopics   = z.fasttopics)
p1 <- ggplot(pdat,aes(x = lfc.mast,y = lfc.fasttopics)) +
  geom_point(color = "mediumblue",shape = 4,size = 0.75) +
  geom_abline(intercept = 0,slope = 1,linetype = "dotted") +
  labs(x = "MAST",y = "fastTopics",title = "LFC estimates") +
  theme_cowplot(font_size = 12)
p2 <- ggplot(pdat,aes(x = lfc.deseq,y = lfc.fasttopics)) +
  geom_point(color = "mediumblue",shape = 4,size = 0.75) +
  geom_abline(intercept = 0,slope = 1,linetype = "dotted") +
  labs(x = "DESeq2",y = "fastTopics",title = "LFC estimates") +
  theme_cowplot(font_size = 12)
p3 <- ggplot(pdat,aes(x = z.deseq,y = z.fasttopics)) +
  geom_point(color = "mediumblue,shape = 4,size = 0.75) +
  geom_abline(intercept = 0,slope = 1,linetype = "dotted") +
  labs(x = "DESeq2",y = "fastTopics",title = "z-scores") +
  theme_cowplot(font_size = 12)
plot_grid(p1,p2,p3,nrow = 1,ncol = 3)
```

```{r lfc-scatterplots-for-paper, echo=FALSE}
ggsave("../plots/mast_deseq2_sims_scatterplots.png",
       plot_grid(p1,p2,p3,nrow = 1,ncol = 3),
       height = 3,width = 9,dpi = 600)
```
