---
title: "Evaluation of fastTopics DE analysis methods in simulations---case of more than 2 topics"
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we evaluate performance of the grade-of-membership DE methods in
simulations, focussing on the case of $K = 6$ topics. A smaller
simulation with $K = 6$ topics was first performed
[here](de_analysis_detailed_look_more.html), and based on this initial
simulation we have conducted a larger set of simulations using the
`run_sims.R` script. Here we summarize the results of this larger set
of simulations.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.show = "hold",fig.align = "center",dpi = 120)
```

Load the packages needed for this analysis, and some additional
functions used to compile the results and generate the plots.

```{r load-pkgs, message=FALSE, warning=FALSE}
library(Matrix)
library(fastTopics)
library(ggplot2)
library(cowplot)
source("../code/de_analysis_functions.R")
```

Load the results of the simulations.

```{r load-sims}
load("../output/sims/sims-k=6.RData")
```

Before comparing the methods, we first assess accuracy of the Monte
Carlo computations by comparing the estimates from two independent
MCMC runs.

```{r mcmc-scatterplots, fig.height=3, fig.width=6}
res["session.info"] <- NULL
pdat <- data.frame(lfc1 = combine_sim_res(res,function (x) c(x$de1$postmean)),
                   lfc2 = combine_sim_res(res,function (x) c(x$de2$postmean)),
                   z1   = combine_sim_res(res,function (x) c(x$de1$z)),
	           z2   = combine_sim_res(res,function (x) c(x$de2$z)))
i <- which(abs(pdat$z1) <= 1 & abs(pdat$z2) <= 1)
j <- which(abs(pdat$z1) > 1 | abs(pdat$z2) > 1)
i <- sample(i,1e5)
pdat <- pdat[c(i,j),]
p1 <- ggplot(pdat,aes(x = lfc1,y = lfc2)) +
  geom_point(color = "dodgerblue",shape = 4,size = 0.75) +
  geom_abline(intercept = 0,slope = 1,linetype = "dotted") +
  labs(x = "first posterior mean",y = "second posterior mean") +
  theme_cowplot(font_size = 12)
p2 <- ggplot(pdat,aes(x = z1,y = z2)) +
  geom_point(color = "dodgerblue",shape = 4,size = 0.75) +
  geom_abline(intercept = 0,slope = 1,linetype = "dotted") +
  labs(x = "first z-score estimate",y = "second z-score estimate") +
  xlim(-22,102) +
  ylim(-22,102) +
  theme_cowplot(font_size = 12)
plot_grid(p1,p2)
```

We see from these scatterplots that the estimates of the posterior
mean LFCs and *z*-scores generated by the two MCMC runs are very
similar.

```{r mcmc-scatterplots-for-paper, echo=FALSE}
ggsave("../plots/mcmc_scatterplots_sims_k=6.png",
       plot_grid(p1,p2),height = 3,width = 6,dpi = 600)
```

Next we consider the K-L divergence measure used in Dey, Hsiao &
Stephens (2017) to rank genes, and compare its ranking to a ranking
based on *p*-values (without using adaptive shrinkage) or *s*-values
(after applying adaptive shrinkage). Since the K-L divergence is not
a signed ranking, we restrict this comparison only to LFCs that are
estimated to be positive.

```{r ranking-histograms, fig.height=2.25, fig.width=7}
get_nonneg_lfcs <- function (de)
  which(de$est > -1e-8)
nonzero_lfc <- combine_sim_res(res,
                 function (x) {
		   n <- nrow(x$dat$F)
		   k <- ncol(x$dat$F)
		   out <- matrix(FALSE,n,k)
                   for (i in 1:n) {
                     y <- x$dat$F[i,]
                     if (max(y) - min(y) > 1e-8) {
                       j <- which.max((y - mean(y))^2)
                       out[i,j] <- TRUE
                     }
		   }
		   return(out[get_nonneg_lfcs(x$de1)])
		 })
noshrink <-
  combine_sim_res(res,function (x) x$de0$lpval[get_nonneg_lfcs(x$de1)])
shrink <-
  combine_sim_res(res,function(x)x$de1$svalue[get_nonneg_lfcs(x$de1)])
lfsr <- combine_sim_res(res,function(x)x$de1$lfsr[get_nonneg_lfcs(x$de1)])
lkl <- combine_sim_res(res,function (x)
         fastTopics:::min_kl_poisson(x$fit$F)[get_nonneg_lfcs(x$de1)])
pdat <- data.frame(nonzero_lfc = factor(nonzero_lfc),
                   noshrink    = 10^(-noshrink),
                   shrink      = shrink,
                   lfsr        = lfsr,
                   lkl         = log10(lkl + 1e-8))
p1 <- ggplot(pdat,aes(x = lkl,color = nonzero_lfc,fill = nonzero_lfc)) +
  geom_histogram(bins = 64,show.legend = FALSE) +
  scale_color_manual(values = c("darkorange","darkblue")) +
  scale_fill_manual(values = c("darkorange","darkblue")) +
  coord_cartesian(ylim = c(0,2.5e4)) +
  labs(x = "log10 K-L divergence",y = "genes",title = "K-L divergence") +
  theme_cowplot(font_size = 12)
p2 <- ggplot(pdat,aes(x = noshrink,color = nonzero_lfc,fill = nonzero_lfc)) +
  geom_histogram(bins = 64,show.legend = FALSE) +
  scale_color_manual(values = c("darkorange","darkblue")) +
  scale_fill_manual(values = c("darkorange","darkblue")) +
  coord_cartesian(ylim = c(0,2.5e4)) +
  labs(x = "p-value",y = "genes",title = "without shrinkage") +
  theme_cowplot(font_size = 12)
p3 <- ggplot(pdat,aes(x = shrink,color = nonzero_lfc,fill = nonzero_lfc)) +
  geom_histogram(bins = 64,show.legend = FALSE) +
  scale_color_manual(values = c("darkorange","darkblue")) +
  scale_fill_manual(values = c("darkorange","darkblue")) +
  coord_cartesian(ylim = c(0,2.5e4)) +
  labs(x = "s-value",y = "genes",title = "with shrinkage") +
  theme_cowplot(font_size = 12)
plot_grid(p1,p2,p3,nrow = 1,ncol = 3)
```

```{r ranking-histograms-for-paper, echo=FALSE}
ggsave("../plots/rankings_sims_k=6.eps",
       plot_grid(p1,p2,p3,nrow = 1,ncol = 3),height = 2.25,width = 7)
```
