---
title: Perform differential expression analysis on droplet topics and clusters
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we perform a differential expression analysis using the topic
model fit to the droplet data, as well as the clusters identified from
this topic model.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the packages used in the analysis below.

```{r load-pkgs, message=FALSE}
library(Matrix)
library(dplyr)
library(fastTopics)
library(tools)
```

Load data and results
---------------------

Load the droplet data, the $k = 7$ Poisson NMF model fit, and the
clusters identified in the [clustering analysis](clusters_droplet.html).

```{r load-data}
load("../data/droplet.RData")
fit <- readRDS("../output/droplet/rds/fit-droplet-scd-ex-k=7.rds")$fit
samples <- readRDS("../output/droplet/clustering-droplet.rds")
```

Perform differential expression analysis
----------------------------------------

Perform differential expression analysis using the multinomial topic
model.

```{r diff-count-analysis-topics}
timing <- system.time(
  diff_count_topics <- diff_count_analysis(fit,counts))
cat(sprintf("Computation took %0.2f seconds.\n",timing["elapsed"]))
```

Next, calculate differential expression statistics using the clusters
that were identified from the topic proportion PCs.

```{r diff-count-analysis-clusters}
fit_clusters <- init_poisson_nmf_from_clustering(counts,samples$cluster)
timing <- system.time(
  diff_count_clusters <- diff_count_analysis(fit_clusters,counts))
cat(sprintf("Computation took %0.2f seconds.\n",timing["elapsed"]))
```

Perform a differential expression analysis after merging the two club
cell topics.

```{r diff-count-analysis-topics-club}
fit_club <- merge_topics(poisson2multinom(fit),c(5,7))
timing <- system.time(
  diff_count_club <- diff_count_analysis(fit_club,counts))
cat(sprintf("Computation took %0.2f seconds.\n",timing["elapsed"]))
```

Finally, to refine differential expression analysis of the basal cells
topic ($k = 2$), perform a differential expression analysis after
removing the "T+N" cluster.

```{r diff-count-analysis-topics-basal}
rows <- which(samples$cluster != "T+N")
fit_basal <- select(poisson2multinom(fit),loadings = rows)
timing <- system.time(
  diff_count_basal <- diff_count_analysis(fit_basal,counts[rows,]))
cat(sprintf("Computation took %0.2f seconds.\n",timing["elapsed"]))
```

Save results
------------

Save the results of the differential expression analyses to an RData
file.

```{r save-results}
save(list = c("diff_count_topics","diff_count_clusters","diff_count_club",
              "diff_count_basal"),
     file = "diff-count-droplet.RData")
resaveRdaFiles("diff-count-droplet.RData")
```
