---
title: Perform differential expression analysis with droplet topics and clusters
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we perform a differential expression analysis using the topic
model fit to the droplet data.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the packages used in the analysis below.

```{r load-pkgs, message=FALSE}
library(Matrix)
library(dplyr)
library(fastTopics)
library(tools)
```

Load data and results
---------------------

Load the droplet data.

```{r load-data}
load("../data/droplet.RData")
```

Load the $k = 7$ Poisson NMF model fit.

```{r load-fit}
fit <- readRDS("../output/droplet/rds/fit-droplet-scd-ex-k=7.rds")$fit
```

Load the clusters identified in the [clustering
analysis](clusters_droplet.html).

```{r load-clusters}
samples <- readRDS("../output/droplet/clustering-droplet.rds")
```

Perform differential expression analysis
----------------------------------------

Perform differential expression analysis using the multinomial topic
model.

```{r diff-count-analysis-topics}
timing <- system.time(
  diff_count_topics <- diff_count_analysis(fit,counts))
cat(sprintf("Computation took %0.2f seconds.\n",timing["elapsed"]))
```

Next, calculate differential expression statistics using the clusters
that were identified from the topic proportion PCs.

```{r diff-count-analysis-clusters}
fit_clusters <- init_poisson_nmf_from_clustering(counts,samples$cluster)
timing <- system.time(
  diff_count_clusters <- diff_count_analysis(fit_clusters,counts))
cat(sprintf("Computation took %0.2f seconds.\n",timing["elapsed"]))
```

Finally, perform one more differential expression analysis after first
merging the two topics for club cells.

```{r diff-count-analysis-topics-merge-club}
fit_merge_club <- merge_topics(poisson2multinom(fit),c(5,7))
timing <- system.time(
  diff_count_merge_club <- diff_count_analysis(fit_merge_club,counts))
cat(sprintf("Computation took %0.2f seconds.\n",timing["elapsed"]))
```

Save results
------------

Save the results of the differential expression analyses to an RData
file.

```{r save-results}
save(list = c("diff_count_topics",
              "diff_count_clusters",
			  "diff_count_merge_club"),
     file = "diff-count-droplet.RData")
resaveRdaFiles("diff-count-droplet.RData")
```
