---
title: Perform differential expression analysis on topics and clusters identified in mixture of FACS-purified PBMC data
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we perform a differential expression analysis using the topic
model fit to the mixture of FACS-purified data, as well as the
clusters identified from this topic model.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the packages used in the analysis below.

```{r load-pkgs, message=FALSE}
library(Matrix)
library(dplyr)
library(fastTopics)
library(tools)
library(ggplot2)
library(ggrepel)
library(cowplot)
source("../code/plots.R")
```

Load data and results
---------------------

Load the UMI count data, the $K = 6$ Poisson NMF model fit, and the
clusters identified in the [clustering analysis](clusters_purified_pbmc.html).

```{r load-data}
load("../data/pbmc_purified.RData")
fit <- readRDS(file.path("../output/pbmc-purified/rds",
                         "fit-pbmc-purified-scd-ex-k=6.rds"))$fit
samples <- readRDS("../output/pbmc-purified/clustering-pbmc-purified.rds")
```

**FOR TESTING ONLY.**

```{r}
rows <- sample(nrow(samples),4700)
counts  <- counts[rows,]
samples <- samples[rows,]
fit     <- select(poisson2multinom(fit),loadings = rows)
```

Perform differential expression analysis
----------------------------------------

Perform differential expression analysis using the multinomial topic
model.

```{r diff-count-analysis-topics}
timing <- system.time(
  diff_count_topics <- diff_count_analysis(fit,counts))
cat(sprintf("Computation took %0.2f seconds.\n",timing["elapsed"]))
```

Next, calculate differential expression statistics using the clusters
that were identified from the topic proportion PCs.

```{r diff-count-analysis-clusters}
fit_clusters <- init_poisson_nmf_from_clustering(counts,samples$cluster)
timing <- system.time(
  diff_count_clusters <- diff_count_analysis(fit_clusters,counts))
cat(sprintf("Computation took %0.2f seconds.\n",timing["elapsed"]))
```

Topics capture shared gene expression patterns
----------------------------------------------

*Add text and code here.*

Save results
------------

Save the results of the differential expression analyses to an RData
file.

```{r save-results}
save(list = c("diff_count_topics","diff_count_clusters"),
     file = "diff-count-pbmc-purified.RData")
resaveRdaFiles("diff-count-pbmc-purified.RData")
```
