---
title: "Examine topic-model-based single-cell likelihoods in FACS-purified PBMC data"
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we calculate single-cell likelihoods to assess how well the
multinomial topic model captures expression in different cells and
cell types.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the packages used in the analysis below.

```{r load-pkgs, message=FALSE}
library(Matrix)
library(fastTopics)
library(ggplot2)
library(cowplot)
source("../code/plots.R")
```

Load the count data, the $K = 6$ topic model fit, and the 7 clusters
identified in the [clustering analysis](clusters_purified_pbmc.html)

```{r load-data}
load("../data/pbmc_purified.RData")
fit <- readRDS(file.path("../output/pbmc-purified/rds",
                         "fit-pbmc-purified-scd-ex-k=6.rds"))$fit
fit <- poisson2multinom(fit)
samples <- readRDS("../output/pbmc-purified/clustering-pbmc-purified.rds")
```

Calculate the multinomial topic model likelihood for each cell.

```{r loglik-1}
loglik <- loglik_multinom_topic_model(counts,fit)
```

This can be used to assess how well the topic model "fits" each cell.

```{r loglik-2, fig.height=2, fig.width=4.5}
pdat <- data.frame(loglik)
p1 <- ggplot(pdat,aes(loglik)) +
  geom_histogram(bins = 64,color = "white",fill = "black") +
  labs(y = "number of cells") +
  theme_cowplot(font_size = 10)
print(p1)
```

Most of the poorly fit cells are in the CD34+ subpopulation:

```{r loglik-3, fig.height=4, fig.width=4}
x <- as.character(samples$celltype)
x[x == "CD4+/CD45RA+/CD25- Naive T"] <- "T cell"
x[x == "CD8+ Cytotoxic T"] <- "T cell"
x[x == "CD4+/CD45RO+ Memory"] <- "T cell"
x[x == "CD8+/CD45RA+ Naive Cytotoxic"] <- "T cell"
x[x == "CD4+ T Helper2"] <- "T cell"
x[x == "CD4+/CD25 T Reg"] <- "T cell"
pdat <- data.frame(loglik = loglik,celltype = x)
p2 <- ggplot(pdat,aes(x = loglik)) +
  geom_histogram(bins = 64,color = "white",fill = "black") +
  facet_grid(rows = vars(celltype),scales = "free") +
  labs(y = "number of cells") +
  theme_cowplot(font_size = 9)
print(p2)
```

```{r save-plots, echo=FALSE, results="hide"}
ggsave("../plots/loglik_histogram_pbmc_purified.eps",
       p1,height = 2,width = 4.5)
ggsave("../plots/loglik_histogram_by_subpop_pbmc_purified.eps",
       p2,height = 2,width = 5.5)
```
