---
title: "Examine topic-model-based single-cell likelihoods in FACS-purified PBMC data"
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we calculate single-cell likelihoods to assess how well the
multinomial topic model captures expression in different cells and
cell types.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the packages used in the analysis below.

```{r load-pkgs, message=FALSE}
library(Matrix)
library(fastTopics)
library(ggplot2)
library(cowplot)
```

Load the count data, the $K = 6$ topic model fit, and the 7 clusters
identified in the [clustering analysis](clusters_purified_pbmc.html).

```{r load-data}
load("../data/pbmc_purified.RData")
fit <- readRDS(file.path("../output/pbmc-purified/rds",
                         "fit-pbmc-purified-scd-ex-k=6.rds"))$fit
fit <- poisson2multinom(fit)
samples <- readRDS("../output/pbmc-purified/clustering-pbmc-purified.rds")
```

Calculate the multinomial topic model likelihood for each cell.

```{r loglik-1}
loglik <- loglik_multinom_topic_model(counts,fit)
```

This can be used to assess how well the topic model "fits" each cell.

```{r loglik-2, fig.height=2, fig.width=5.5}
pdat <- data.frame(loglik)
p1 <- ggplot(pdat,aes(loglik)) +
  geom_histogram(bins = 64,color = "white",fill = "black") +
  scale_x_continuous(breaks = seq(-20000,0,2500)) +
  labs(y = "number of cells") +
  theme_cowplot(font_size = 10)
print(p1)
```

Most of the poorly fit cells are in the CD34+ subpopulation:

```{r loglik-3, fig.height=4, fig.width=6}
x <- as.character(samples$celltype)
x[x == "CD14+ Monocyte"] <- "CD14+"
x[x == "CD4+/CD45RA+/CD25- Naive T"] <- "T cell"
x[x == "CD4+/CD45RO+ Memory"] <- "T cell"
x[x == "CD8+/CD45RA+ Naive Cytotoxic"] <- "T cell"
x[x == "CD4+ T Helper2"] <- "T cell"
x[x == "CD4+/CD25 T Reg"] <- "T cell"
x <- factor(x)
pdat <- data.frame(loglik = loglik,celltype = x)
p2 <- ggplot(pdat,aes(x = loglik)) +
  geom_histogram(bins = 64,color = "white",fill = "black") +
  facet_wrap(vars(celltype),scales = "free_y",ncol = 2) +
  scale_x_continuous(breaks = seq(-20000,0,2500)) +
  labs(y = "number of cells") +
  theme_cowplot(font_size = 9)
print(p2)
```

Here, we compare the single-cell likelihoods under the multinomial
topic model against the likelihoods under a simple "hard clustering"
model in which the cells in each FACS subpopulation share the same
underlying pattern of expression. This serves as a sanity check, as we
expect the fit to improve with the more flexible topic model.

```{r loglik-4}
fit_subpop <- init_poisson_nmf_from_clustering(counts,x)
fit_subpop <- poisson2multinom(fit_subpop)
loglik_subpop <- loglik_multinom_topic_model(counts,fit_subpop)
```

Indeed, the topic model provides a better fit for almost all cells:

```{r loglik-5, fig.width=4, fig.height=2.5}
facs_colors <- c("dodgerblue",  # B cells
                 "forestgreen", # CD14+
                 "darkmagenta", # CD34+
                 "gray",        # NK cells
                 "tomato",      # cytotoxic T cells
                 "gold")        # T cells
pdat <- data.frame(x = loglik_subpop,y = loglik,celltype = x)
p3 <- ggplot(pdat,aes(x = x,y = y,fill = celltype)) +
  geom_point(shape = 21,color = "white") +
  geom_abline(intercept = 0,slope = 1,linetype = "dotted") +
  scale_x_continuous(limits = c(-15000,0),breaks = seq(-20000,0,2500)) +
  scale_y_continuous(limits = c(-15000,0),breaks = seq(-20000,0,2500)) +
  scale_fill_manual(values = facs_colors) +
  labs(x = "cluster model",y = "topic model",fill = "FACS subpopulation") +
  theme_cowplot(font_size = 9)
print(p3)
```

The improvement in fit is greatest for cells in the CD34+ and T cell
FACS subpopulations:

```{r loglik-6, fig.width=6.5, fig.height=3.5}
p4 <- ggplot(pdat,aes(x = x,y = y,fill = celltype)) +
  geom_point(shape = 21,color = "white") +
  geom_abline(intercept = 0,slope = 1,linetype = "dotted") +
  facet_wrap(vars(celltype)) +
  scale_x_continuous(limits = c(-15000,0),breaks = seq(-20000,0,5000)) +
  scale_y_continuous(limits = c(-15000,0),breaks = seq(-20000,0,5000)) +
  scale_fill_manual(values = facs_colors) +
  labs(x = "cluster model",y = "topic model",fill = "FACS subpopulation") +
  theme_cowplot(font_size = 9)
print(p4)
```

```{r save-plots, echo=FALSE, results="hide"}
ggsave("../plots/loglik_histogram_pbmc_purified.eps",
       p1,height = 2,width = 4.5)
ggsave("../plots/loglik_histogram_by_subpop_pbmc_purified.eps",
       p2,height = 4,width = 4.5)
ggsave("../plots/loglik_scatterplot_all_pbmc_purified.png",
       p3,height = 3.5,width = 5)
ggsave("../plots/loglik_scatterplot_by_subpop_pbmc_purified.png",
       p4,height=4,width = 7.5)
```
