---
title: Visualize and interpret topics in PBMC data
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we look closely, and compare, the topic modeling results for the
two closely related data sets from [Zheng *et al* (2017)][zheng-2017],
the mixture of FACS-purified PBMC data and the "unsorted" 68k PBMC
data. The goal is to illustrate how the topic models fitted to these
data sets can be used to learn about the structure in the data,
identify clusters, and interpret the clusters and topics as "cell
types" or "gene programs".

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the packages used in the analysis below, as well as additional
functions that will be used to generate some of the plots.

```{r load-pkgs, message=FALSE}
library(dplyr)
library(fastTopics)
library(ggplot2)
library(cowplot)
source("../code/plots.R")
```

Mixture of FACS-purified PBMC data
----------------------------------

We begin with the mixture of FACS-purified PBMC data. Note that the
count data are no longer needed at this stage.

```{r load-purified-data}
load("../data/pbmc_purified.RData")
samples_purified <- samples
rm(samples,genes,counts)
```

Load the $k = 6$ Poisson NMF model fit.

```{r load-purified-fit}
fit_purified <-
  readRDS("../output/pbmc-purified/rds/fit-pbmc-purified-scd-ex-k=6.rds")$fit
```

Here we explore the structure of the single-cell data as inferred by
the topic model. Specifically, we will use PCA to uncover structure in
the topic proportions. Although PCA is simple, we will see that it
works well, and avoids the complications of the popular t-SNE and UMAP
nonlinear dimensionality reduction methods.

```{r pbmc-purified-clustering-1}
fit <- poisson2multinom(fit_purified)
pca <- prcomp(fit$L)$x
```

Three large clusters are evident from first two PCs (there is also
finer-scale structure which we will examine below). We label these
clusters as "A", "B" and "C".

```{r pbmc-purified-clustering-2, fig.width=5, fig.height=4}
n   <- nrow(pca)
x   <- rep("C",n)
pc1 <- pca[,"PC1"]
pc2 <- pca[,"PC2"]
x[pc1 + 0.2 > pc2] <- "A"
x[pc2 > 0.25] <- "B"
x[(pc1 + 0.4)^2 + (pc2 + 0.1)^2 < 0.07] <- "C"
samples_purified$cluster <- x
p1 <- pca_plot_with_labels(fit_purified,c("PC1","PC2"),
                           samples_purified$cluster) +
      labs(fill = "cluster")
print(p1)
```

Most of the samples are in cluster A:

```{r pbmc-purified-clustering-3}
table(x)
```

Within cluster C there are two fairly well-defined subclusters
(labeled "C1" and "C2"). There are perhaps other, less defined
subclusters that are less defined, but in this analysis we focus on
the largest, most obvious clusters.

```{r pbmc-purified-clustering-4, fig.width=5, fig.height=4}
rows <- which(samples_purified$cluster == "C")
fit  <- select(poisson2multinom(fit_purified),loadings = rows)
pca  <- prcomp(fit$L)$x
n    <- nrow(pca)
x    <- rep("C3",n)
pc1  <- pca[,1]
pc2  <- pca[,2]
x[pc1 < 0 & pc2 < 0.4] <- "C1"
x[pc1 > 0.5 & pc2 < 0.3] <- "C2"
samples_purified[rows,"cluster"] <- x
p2 <- pca_plot_with_labels(fit,c("PC1","PC2"),x) +
      labs(fill = "cluster")
print(p2)
```

The two subclusters, C1 and C2, account for most of the samples in
cluster C:

```{r pbmc-purified-clustering-5}
table(x)
```

Now we turn to cluster A. Within this cluster, there is a large
subcluster, which we label as "A1". (This cluster is much less
distinct than the other clusters we have seen so far, and may not show
up clearly in this plot---you may need to zoom in on the plot to see
the clustering.) Otherwise, there are is obvious additional clustering
of the samples within cluster A.

```{r pbmc-purified-clustering-6, fig.width=5, fig.height=4}
rows <- which(samples_purified$cluster == "A")
fit  <- select(poisson2multinom(fit_purified),loadings = rows)
pca  <- prcomp(fit$L)$x
n    <- nrow(fit$L)
x    <- rep("A2",n)
pc1  <- pca[,1]
pc2  <- pca[,2]
x[pc1 > 0.58 - pc2 | pc1 > 0.7] <- "A1"
samples_purified[rows,"cluster"] <- x
p3 <- pca_plot_with_labels(fit,c("PC1","PC2"),x) +
      labs(fill = "cluster")
print(p3)
```

In summary, we have subdivided the data into 6 clusters:

```{r pbmc-purified-clustering-7}
samples_purified$cluster <- factor(samples_purified$cluster)
table(samples_purified$cluster)
```

The structure plot provides an evocative visual summary of topic
proportions in each of these six clusters:

```{r pbmc-purified-structure-plot, fig.width=7, fig.height=1.5}
set.seed(1)
pbmc_purified_topic_colors <- c("gold","forestgreen","dodgerblue",
                                "gray","greenyellow","magenta")
pbmc_purified_topics <- c(2,5,3,1,4,6)
rows <- sort(c(sample(which(samples_purified$cluster == "A1"),250),
               sample(which(samples_purified$cluster == "A2"),1200),
               sample(which(samples_purified$cluster == "B"),250),
               sample(which(samples_purified$cluster == "C1"),250),
               sample(which(samples_purified$cluster == "C2"),200),
               sample(which(samples_purified$cluster == "C3"),200)))
p4 <- structure_plot(select(poisson2multinom(fit_purified),loadings = rows),
                     grouping = samples_purified[rows,"cluster"],
                     topics = pbmc_purified_topics,
                     colors = pbmc_purified_topic_colors[pbmc_purified_topics],
                     gap = 40,num_threads = 4)
print(p4)
```

Out of the 6 topics, 4 of them ($k = 2, 3, 4, 5$) align closely with
the clusters (labeled A1, B, C1, C2). And, indeed, they align closely
with their inclusion in the indivvidual FACS-purified data sets:

```{r pbmc-purified-clusters-vs-celltypes}
with(samples_purified,table(celltype,cluster))
```

For example, cluster B corresponds almost perfectly to the B-cell data
set, and the largest cluster---cluster A2---is comprised of the T-cell
data sets. It is also interesting that many of the samples labeled as
"CD34+" are not assigned to the CD34+ cluster (C1), which probably
reflects the fact that the this population was much less pure (45%)
than the others, and so probably contained other cell types.

Cluster C3 is a heterogeneous cluster that we do not investigate
further. Cluster A2---see also the PCA plot above---is a clear example
where topic modeling is more appropriate than clustering because any
additional clustering of the data will be arbitrary.

Unsorted 68k PBMC data
----------------------

Next, we turn to the 68k data set.

```{r load-68k-data}
load("../data/pbmc_68k.RData")
samples_68k <- samples
rm(samples,genes,counts)
```

Load the $k = 6$ Poisson NMF model fit.

```{r load-68k-fit}
fit_68k <- readRDS("../output/pbmc-68k/rds/fit-pbmc-68k-scd-ex-k=6.rds")$fit
```

```{r pbmc-68k-clustering-1}
fit <- poisson2multinom(fit_68k)
pca <- prcomp(fit$L)$x
```

In this case, we find least three distinct clusters in the projection
onto PCs 3 and 4. We label these clusters "A", "B" and "C", as above,
but this labeling does not imply a connection between the two sets
of clusters.

```{r pbmc-68k-clustering-2, fig.width=5, fig.height=4}
n <- nrow(pca)
x <- rep("A",n)
pc3 <- pca[,"PC3"]
pc4 <- pca[,"PC4"]
x[pc4 < -0.13 | pc3/1.9 - 0.17 > pc4] <- "B"
x[pc4 < -0.75] <- "C"
samples_68k$cluster <- x
p5 <- pca_plot_with_labels(fit_68k,c("PC3","PC4"),x) +
      labs(fill = "cluster")
print(p5)
```

The vast majority of the cells are in cluster A:

```{r pbmc-68k-clustering-3}
table(samples_68k$cluster)
```

Looking more closely at cluster B:

```{r pbmc-68k-clustering-4, fig.width=5, fig.height=4}
rows <- which(samples_68k$cluster == "B")
fit  <- select(poisson2multinom(fit_68k),loadings = rows)
pca  <- prcomp(fit$L)$x
n    <- nrow(pca)
x    <- rep("B3",n)
pc1  <- pca[,"PC1"]
x[pc1 > -0.05] <- "B1"
x[pc1 < -0.3] <- "B2"
samples_68k[rows,"cluster"] <- x
p6 <- pca_plot_with_labels(fit,c("PC1","PC2"),x) +
      labs(fill = "cluster")
print(p6)
```

Looking more closely at cluster A:

```{r pbmc-68k-clustering-5, fig.width=5, fig.height=4}
rows <- which(samples_68k$cluster == "A")
fit  <- select(poisson2multinom(fit_68k),loadings = rows)
pca  <- prcomp(fit$L)$x
n    <- nrow(pca)
x    <- rep("A3",n)
pc2  <- pca[,"PC2"]
pc3  <- pca[,"PC3"]
x[2.5*pc3 < 0.3 - pc2] <- "A1"
x[pc3 > 0.8 - pc2] <- "A2"
samples_68k[rows,"cluster"] <- x
p7 <- pca_plot_with_labels(fit,c("PC2","PC3"),x) +
      labs(fill = "cluster")
print(p7)
```

Within cluster A, the vast majority of the samples are assigned to the
A1 subcluster:

```{r pbmc-68k-clustering-6}
table(x)
```

In summary, we have subdivided these data into 7 clusters:

```{r pbmc-68k-clustering-7}
samples_68k$cluster <- factor(samples_68k$cluster)
table(samples_68k$cluster)
```

**TO DO:** Add text here.

```{r pbmc-68k-structure-plot, fig.width=7, fig.height=1.5}
set.seed(1)
pbmc_68k_topic_colors <- c("gold","lightskyblue","dodgerblue",
                           "gray","forestgreen","magenta")
pbmc_68k_topics <- c(2,5,1,3,4,6)
rows <- sort(c(sample(which(samples_68k$cluster == "A1"),1200),
               sample(which(samples_68k$cluster == "A2"),500),
               sample(which(samples_68k$cluster == "A3"),300),
               sample(which(samples_68k$cluster == "B1"),500),
               sample(which(samples_68k$cluster == "B2"),300),
               which(samples_68k$cluster == "B3"),
               which(samples_68k$cluster == "C")))
p8 <- structure_plot(select(poisson2multinom(fit_68k),loadings = rows),
                     grouping = samples_68k[rows,"cluster"],
                     topics = pbmc_68k_topics,
                     colors = pbmc_68k_topic_colors[pbmc_68k_topics],
                     gap = 40,num_threads = 4,verbose = FALSE)
print(p8)
```

Comparison to Zheng et al (2017) cell-type labeling:

```{r pbmc-68k-clusters-vs-celltypes}
with(samples_68k,table(celltype,cluster))
```

[zheng-2017]: https://doi.org/10.1038/ncomms14049
